Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-13 talk/pepm2011/paper.tex
make the code nicer

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}

@@

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through it.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \texttt{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \texttt{if} statement is true, because if
it isn't, the rest of the trace is not valid.

@@

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type. The objects support only
two operations, \texttt{add}, which adds two objects (promoting ints to floats in a
mixed addition) and \texttt{is\_positive}, which returns whether the number is greater
than zero. The implementation of \texttt{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
\ignore{
class Base(object):
    def add(self, other):
        """ add self to other """
        raise NotImplementedError("abstract base")
    def add__int(self, intother):
        """ add intother to self,
            where intother is an integer """
        raise NotImplementedError("abstract base")
    def add__float(self, floatother):
        """ add floatother to self,
            where floatother is a float """
        raise NotImplementedError("abstract base")
    def is_positive(self):
        """ returns whether self is positive """
        raise NotImplementedError("abstract base")
        }
\begin{figure}
\begin{verbatim}
class Base(object):

class BoxedInteger(Base):
    def __init__(self, intval):
        self.intval = intval
    def add(self, other):
        return other.add__int(self.intval)
    def add__int(self, intother):
        return BoxedInteger(intother + self.intval)
    def add__float(self, floatother):
        floatvalue = floatother + float(self.intval)
        return BoxedFloat(floatvalue)
    def is_positive(self):
        return self.intval > 0

class BoxedFloat(Base):
    def __init__(self, floatval):
        self.floatval = floatval
    def add(self, other):
        return other.add__float(self.floatval)
    def add__int(self, intother):
        floatvalue = float(intother) + self.floatval
        return BoxedFloat(floatvalue)
    def add__float(self, floatother):
        return BoxedFloat(floatother + self.floatval)
    def is_positive(self):
        return self.floatval > 0.0

def f(y):
    res = BoxedInteger(0)
    while y.is_positive():
        res = res.add(y).add(BoxedInteger(-100))
        y = y.add(BoxedInteger(-1))
    return res
\end{verbatim}
\caption{An ``interpreter'' for a tiny Dynamic Language written in RPython}
%\caption{A Simple Object Model and an Example Function Using it}
\label{fig:objmodel}
\end{figure}

Using these classes to implement arithmetic shows the basic problem that a
dynamic language implementation has. All the numbers are instances of either
\texttt{BoxedInteger} or \texttt{BoxedFloat}, thus they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
thus putting pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

To understand the problems more directly, let us consider the simple
interpreter function
\texttt{f} that uses the object model (see the bottom of
Figure~\ref{fig:objmodel}).

@@

The loop in \texttt{f} iterates \texttt{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \texttt{is\_positive} and even two for each
call to \texttt{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \texttt{f} is simply a sequence of
float or integer additions.

\begin{figure}
\texttt{
\begin{tabular}{l}
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace} (lines starting with the hash ``\#'' are comments).

@@

The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \texttt{y} and \texttt{res} in the original function.

XXX explain set and get + int_add briefly

The trace shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new} (corresponding to object creation),
\texttt{set/get} (corresponding to attribute reads/writes) and
\texttt{guard\_class} operations (corresponding to method calls).
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.

Note how the functions that are called by \texttt{f} are automatically inlined
into the trace. The method calls are always preceded by a \texttt{guard\_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\texttt{guard\_class} performs a precise
class check, not checking for subclasses.} These guards make the trace specific
to the situation where \texttt{y} is really a \texttt{BoxedInteger}, it can
already be said to be specialized for \texttt{BoxedIntegers}. When the trace is
turned into machine code and then executed with \texttt{BoxedFloats}, the
first \texttt{guard\_class} instruction will fail and execution will continue
using the interpreter.

@@

Some of the operations within this trace are \texttt{new} operations, which each
create a new instance of some class. These instances are used for a while, e.g.
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a
non-inlined function via a residual call.

Together with the \texttt{new} operations, the figure shows the lifetimes of the
created objects. The objects that are created within a trace using \texttt{new}
fall into one of several categories:

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check the guard does is statically known.

@@

\texttt{
\begin{tabular}{l}
$p_{5}$ = new(BoxedInteger) \\
set($p_{5}$, intval, $i_{4}$) \\
$p_{6}$ = new(BoxedInteger) \\
set($p_{6}$, intval, -100) \\
\end{tabular}
}

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
whose \texttt{intval} field contains the constant -100.

@@

\texttt{
\begin{tabular}{l}
guard\_class($p_{5}$, BoxedInteger) \\
$i_{7}$ = get($p_{5}$, intval) \\
\# inside BoxedInteger.add \\
guard\_class($p_{6}$, BoxedInteger) \\
\# inside BoxedInteger.add\_\_int \\
$i_{8}$ = get($p_{6}$, intval) \\
$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
\end{tabular}
}

The \texttt{guard\_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

\texttt{
\begin{tabular}{l}
$i_{9}$ = int\_add($i_{4}$, -100) \\
\end{tabular}
}

@@

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively) This means that instead of the jump,
the following operations are emitted:

\texttt{
\begin{tabular}{l}
$p_{15}$ = new(BoxedInteger) \\
set($p_{15}$, intval, $i_{14}$) \\
$p_{10}$ = new(BoxedInteger) \\
set($p_{10}$, intval, $i_{9}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{get} operations
and \texttt{guard\_class} operations).

@@

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \texttt{guard\_class} operations, from the original seven.

@@

Traces are lists of operations. The operations considered here are \texttt{new} (to make
a new object), \texttt{get} (to read a field out of an object), \texttt{set} (to write a field
into an object) and \texttt{guard\_class} (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are initialized to a real
location before being read, otherwise the trace is malformed.

@@

The \texttt{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
heap under a fresh location $l$ and adds the result variable to the environment,
mapping it to the new location $l$.

The \texttt{get} operation reads a field $F$ out of an object, and adds the result
variable to the environment, mapping it to the read location. The heap is
unchanged.

The \texttt{set} operation changes field $F$ of an object stored at the location that
variable $v$ maps to. The new value of the field is the location in variable
$u$. The environment is unchanged.

The \texttt{guard\_class} operation is used to check whether the object stored at the location
that variable $v$ maps to is of type $T$. If that is the case, then execution
continues without changing heap and environment. Otherwise, execution is
stopped.

@@

When the optimizer sees a \texttt{new} operation, it optimistically removes it and
assumes that the resulting object can stay static. The optimization for all
further operations is split into two cases. One case is for when the
involved variables are in the static heap, which means that the operation can be
performed at optimization time and removed from the trace. These rules mirror
the execution semantics closely. The other case is for when not enough is known about
the variables, and the operation has to be residualized.

If the argument $v$ of a \texttt{get} operation is mapped to something in the static
heap, the \texttt{get} can be performed at optimization time. Otherwise, the \texttt{get}
operation needs to be residualized.

If the first argument $v$ to a \texttt{set} operation is mapped to something in the
static heap, then the \texttt{set} can performed at optimization time and the static heap
updated. Otherwise the \texttt{set} operation needs to be residualized. This needs to be
done carefully, because the new value for the field, from the variable $u$,
could itself be static, in which case it needs to be lifted first.

If a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
can be performed at optimization time, which means the operation can be removed
if the types match. If the type check fails statically or if the object is not
in the static heap, the \texttt{guard\_class} is residualized. This also needs to
lift the variable on which the \texttt{guard\_class} is performed.

@@

Lifting a static object needs to recursively lift its fields. Some care needs to
be taken when lifting a static object, because the structures described by the
static heap can be cyclic. To make sure that the same static object is not lifted
twice, the \texttt{liftfield} operation removes it from the static heap \emph{before}
recursively lifting its fields.

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

\usepackage{listings}

\usepackage[T1]{fontenc}
\usepackage{beramono}

\definecolor{gray}{rgb}{0.3,0.3,0.3}

\lstset{
  basicstyle=\setstretch{1.1}\ttfamily\footnotesize,
  language=Python,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{gray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  keywords={def,while,if,elif,return,class,get,set,new,guard_class}
}

@@

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through it.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \lstinline{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \lstinline{if} statement is true, because if
it isn't, the rest of the trace is not valid.

@@

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type. The objects support only
two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition) and \lstinline{is_positive}, which returns whether the number is greater
than zero. The implementation of \lstinline{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
\ignore{
class Base(object):
    def add(self, other):
        """ add self to other """
        raise NotImplementedError("abstract base")
    def add__int(self, intother):
        """ add intother to self,
            where intother is an integer """
        raise NotImplementedError("abstract base")
    def add__float(self, floatother):
        """ add floatother to self,
            where floatother is a float """
        raise NotImplementedError("abstract base")
    def is_positive(self):
        """ returns whether self is positive """
        raise NotImplementedError("abstract base")
        }
\begin{figure}
\begin{lstlisting}[mathescape]
class Base(object):
      ...

class BoxedInteger(Base):
   def __init__(self, intval):
      self.intval = intval

   def add(self, other):
      return other.add__int(self.intval)

   def add__int(self, intother):
      return BoxedInteger(intother + self.intval)

   def add__float(self, floatother):
      floatvalue = floatother + float(self.intval)
      return BoxedFloat(floatvalue)

   def is_positive(self):
      return self.intval > 0

class BoxedFloat(Base):
   def __init__(self, floatval):
      self.floatval = floatval

   def add(self, other):
      return other.add__float(self.floatval)

   def add__int(self, intother):
      floatvalue = float(intother) + self.floatval
      return BoxedFloat(floatvalue)

   def add__float(self, floatother):
      return BoxedFloat(floatother + self.floatval)

   def is_positive(self):
      return self.floatval > 0.0

def f(y):
   res = BoxedInteger(0)
   while y.is_positive():
      res = res.add(y).add(BoxedInteger(-100))
      y = y.add(BoxedInteger(-1))
   return res
\end{lstlisting}
\caption{An ``interpreter'' for a tiny Dynamic Language written in RPython}
%\caption{A Simple Object Model and an Example Function Using it}
\label{fig:objmodel}
\end{figure}

Using these classes to implement arithmetic shows the basic problem that a
dynamic language implementation has. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, thus they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
thus putting pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

To understand the problems more directly, let us consider the simple
interpreter function
\lstinline{f} that uses the object model (see the bottom of
Figure~\ref{fig:objmodel}).

@@

The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \lstinline{is_positive} and even two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions.

\begin{figure}
\begin{lstlisting}[mathescape]
# arguments to the trace: $p_{0}$, $p_{1}$
# inside f: res.add(y)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)

# inside f: BoxedInteger(-100) 
$p_{6}$ = new(BoxedInteger)
    # inside BoxedInteger.__init__
    set($p_{6}$, intval, -100)

# inside f: .add(BoxedInteger(-100))
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{7}$ = get($p_{5}$, intval)
    guard_class($p_{6}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{8}$ = get($p_{6}$, intval)
        $i_{9}$ = int_add($i_{7}$, $i_{8}$)
        $p_{10}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{10}$, intval, $i_{9}$)

# inside f: BoxedInteger(-1)
$p_{11}$ = new(BoxedInteger)
    # inside BoxedInteger.__init__
    set($p_{11}$, intval, -1)

# inside f: y.add(BoxedInteger(-1))
guard_class($p_{0}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{12}$ = get($p_{0}$, intval)
    guard_class($p_{11}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{13}$ = get($p_{11}$, intval)
        $i_{14}$ = int_add($i_{12}$, $i_{13}$)
        $p_{15}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{15}$, intval, $i_{14}$)

# inside f: y.is_positive()
guard_class($p_{15}$, BoxedInteger)
    # inside BoxedInteger.is_positive
    $i_{16}$ = get($p_{15}$, intval)
    $i_{17}$ = int_gt($i_{16}$, 0)
# inside f
guard_true($i_{17}$)
jump($p_{15}$, $p_{10}$)
\end{lstlisting}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

If the function is executed using the tracing JIT, with \lstinline{y} being a
\lstinline{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace} (lines starting with the hash ``\#'' are comments).

@@

The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \lstinline{y} and \lstinline{res} in the original function.

XXX explain set and get + int\_add briefly

The trace shows the inefficiencies of \lstinline{f} clearly, if one
looks at the number of \lstinline{new} (corresponding to object creation),
\lstinline{set/get} (corresponding to attribute reads/writes) and
\lstinline{guard_class} operations (corresponding to method calls).
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.

Note how the functions that are called by \lstinline{f} are automatically inlined
into the trace. The method calls are always preceded by a \lstinline{guard_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\lstinline{guard_class} performs a precise
class check, not checking for subclasses.} These guards make the trace specific
to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}, it can
already be said to be specialized for \lstinline{BoxedIntegers}. When the trace is
turned into machine code and then executed with \lstinline{BoxedFloats}, the
first \lstinline{guard_class} instruction will fail and execution will continue
using the interpreter.

@@

Some of the operations within this trace are \lstinline{new} operations, which each
create a new instance of some class. These instances are used for a while, e.g.
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a
non-inlined function via a residual call.

Together with the \lstinline{new} operations, the figure shows the lifetimes of the
created objects. The objects that are created within a trace using \lstinline{new}
fall into one of several categories:

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \lstinline{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \lstinline{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \lstinline{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\lstinline{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check the guard does is statically known.

@@

\begin{lstlisting}[mathescape,xleftmargin=20pt]
$p_{5}$ = new(BoxedInteger)
set($p_{5}$, intval, $i_{4}$)
$p_{6}$ = new(BoxedInteger)
set($p_{6}$, intval, -100)
\end{lstlisting}

The static object associated with $p_{5}$ would know that it is a
\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \lstinline{BoxedInteger}
whose \lstinline{intval} field contains the constant -100.

@@

\begin{lstlisting}[mathescape,xleftmargin=20pt]
guard_class($p_{5}$, BoxedInteger)
$i_{7}$ = get($p_{5}$, intval)
# inside BoxedInteger.add
guard_class($p_{6}$, BoxedInteger)
# inside BoxedInteger.add__int
$i_{8}$ = get($p_{6}$, intval)
$i_{9}$ = int_add($i_{7}$, $i_{8}$)
\end{lstlisting}

The \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \lstinline{BoxedInteger}. The \lstinline{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

\begin{lstlisting}[mathescape,xleftmargin=20pt]
$i_{9}$ = int_add($i_{4}$, -100)
\end{lstlisting}

@@

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively) This means that instead of the jump,
the following operations are emitted:

\begin{lstlisting}[mathescape,xleftmargin=20pt]
$p_{15}$ = new(BoxedInteger)
set($p_{15}$, intval, $i_{14}$)
$p_{10}$ = new(BoxedInteger)
set($p_{10}$, intval, $i_{9}$)
jump($p_{15}$, $p_{10}$)
\end{lstlisting}

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \lstinline{get} operations
and \lstinline{guard_class} operations).

@@

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \lstinline{guard_class} operations, from the original seven.

@@

Traces are lists of operations. The operations considered here are \lstinline{new} (to make
a new object), \lstinline{get} (to read a field out of an object), \lstinline{set} (to write a field
into an object) and \lstinline{guard_class} (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are initialized to a real
location before being read, otherwise the trace is malformed.

@@

The \lstinline{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
heap under a fresh location $l$ and adds the result variable to the environment,
mapping it to the new location $l$.

The \lstinline{get} operation reads a field $F$ out of an object, and adds the result
variable to the environment, mapping it to the read location. The heap is
unchanged.

The \lstinline{set} operation changes field $F$ of an object stored at the location that
variable $v$ maps to. The new value of the field is the location in variable
$u$. The environment is unchanged.

The \lstinline{guard_class} operation is used to check whether the object stored at the location
that variable $v$ maps to is of type $T$. If that is the case, then execution
continues without changing heap and environment. Otherwise, execution is
stopped.

@@

When the optimizer sees a \lstinline{new} operation, it optimistically removes it and
assumes that the resulting object can stay static. The optimization for all
further operations is split into two cases. One case is for when the
involved variables are in the static heap, which means that the operation can be
performed at optimization time and removed from the trace. These rules mirror
the execution semantics closely. The other case is for when not enough is known about
the variables, and the operation has to be residualized.

If the argument $v$ of a \lstinline{get} operation is mapped to something in the static
heap, the \lstinline{get} can be performed at optimization time. Otherwise, the \lstinline{get}
operation needs to be residualized.

If the first argument $v$ to a \lstinline{set} operation is mapped to something in the
static heap, then the \lstinline{set} can performed at optimization time and the static heap
updated. Otherwise the \lstinline{set} operation needs to be residualized. This needs to be
done carefully, because the new value for the field, from the variable $u$,
could itself be static, in which case it needs to be lifted first.

If a \lstinline{guard_class} is performed on a variable that is in the static heap, the type check
can be performed at optimization time, which means the operation can be removed
if the types match. If the type check fails statically or if the object is not
in the static heap, the \lstinline{guard_class} is residualized. This also needs to
lift the variable on which the \lstinline{guard_class} is performed.

@@

Lifting a static object needs to recursively lift its fields. Some care needs to
be taken when lifting a static object, because the structures described by the
static heap can be cyclic. To make sure that the same static object is not lifted
twice, the \lstinline{liftfield} operation removes it from the static heap \emph{before}
recursively lifting its fields.
